Effect of Black Hole Spins on Christodoulou Memory in Binary Mergers and its implication on Observing Gravitational Wave Bursts With Memory in Pulsar Timing Array

### 2019-10-09 notes/Paper.tex
updated paper

The growth in the nonlinear memory signal amplitude from super massive black hole occurs on the time scale of a few days so the effect on timing residuals can be seen if observed over a few weeks period of time. A method for observing Nonlinear memory signal from super massive black hole mergers using pulsar timing has been proposed in ref\cite{Pshirkov2010}. The authors have shown that the present day pulsar timing precision allows one to detect BWM due to  SMBH merger at distance upto 1 Gpc (for case of equal mass $10^{8}$ solar mass). The details regarding the calculation of the timing residuals can be seen in \cite{Pshirkov2010}, where they consider the memory event as a step function which occurs in the middle of an observation time of 10 years. The method described in \cite{Pshirkov2010} is general and can be applied to case where the memory jump is not at step function. In our analysis we explore how the nonlinear memory grows during two weeks interval to time for different ranges of mass and spin parameter values. We also calculate the timing residuals as done in \cite{Pshirkov2010} and explore how it changes as a function of total mass and spin of black holes as shown in Fig[\ref{fig:MemoryMassVsSpin}].
\par 
Future observations with increased precision would be able to look at the memory growth on even shorter time of two weeks. The memory growth on the time scale of two weeks does not look like a step function, instead it's shape depends upon the spins of the merging black holes. In Fig[\ref{fig:MemoryGrowthTwoweeks}].we show the memory growth during two weeks interval of time with markers on complete memory during for different range of spin and total mass of the binary. The solid line show the complete memory growth in growth in geometrical units. It can be seen the as the total mass of binary becomes larger the two weeks interval of time shrinks in geometrical units. We have chosen this two weeks interval of time such that the change in memory during this time interval is maximum. For the pulsar timing observation we compute the residuals during this two week interval of time following the method described in \cite{Pshirkov2010}. It can be seen in Fig[\ref{fig:MemoryGrowthTwoweeks}] that the memory growth during this two weeks interval of time look very for different mass values. For higher masses ($10^{10}M_{\odot}$), this looks like a straight line. For lower masses this looks like a nonlinear growth. In Fig[\ref{fig:resedualG}] we show the residual before(pre-fit) and after  (post-fit) a quadratic subtraction. It can be seen that the residual growth fits better to a pre-fit residuals 
%%\begin{widetext}
\begin{figure}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksGeoUnits.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow8MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow9MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow11MSun.pdf}
	\caption{The top figure shows the memory growth during merger of a binary with time measured in geometric units with total spin magnitude as indicated above. The two weeks interval of time for a given mass as indicated above. The bottom three figures show the the memory growth in days for Mass $10^{8}M_{\odot}$, $10^{9}M_{\odot}$ and $10^{11}M_{\odot}$ respectively.}
	\label{fig:resedualG}
\end{figure} 
%\end{widetext}

@@

\begin{figure}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDays10pow8MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDays10pow8MSunQuadSub.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDays10pow10MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDays10pow10MSunQuadSub.pdf}
	\caption{The figure above shows residual growth in a given interval of time. The top left plot shows this residual growth during two weeks interval of time for different spin parameter values and the top right plot shows the residual growth after subtracting a quadratic fit for maximally spinning case and a binary with $M = 10^{8}\, M_{\odot}$. The bottom figure shows the similar plot for a binary with $M = 10^{11}\, M_{\odot}$.}
	\label{fig:resedualG}
\end{figure}

\begin{figure}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDaysPreFit.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDaysPostfit.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/ResidualGrowth5YearsDiffSpinsInDaysPrefit.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/ResidualGrowth5YearsDiffSpinsInDaysPostfit.pdf}
	\caption{The figure above shows residual growth in a given interval of time. The top left plot shows this residual growth during two weeks interval of time and the top right plot shows the residual growth after subtracting a quadratic fit. The bottom figure shows the similar plot for 5 years interval of time.}
	\label{fig:resedualGrowth}
\end{figure}

\begin{figure}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/MemoryResSpinvsMass14Days.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/MemoryResSpinvsMass1825Days.pdf}
	\caption{The top figure shows the root mean square of post-fit timing residuals growth as a function of Mass and Spin magnitude $\chi$. The left plot shows this post-fit residual fit during two weeks interval of time and right plot shows this for 5 years interval.}
	\label{fig:MemoryMassVsSpin}
\end{figure}
\end{widetext}
\section{CONCLUSIONS}
Observation of Christodoulou memory would be an important test of general relativity in strong field regime. Effort to detect this memory effect in low frequency band from super massive black holes have been undertaken by pulsar timing array experiments \cite{Pshirkov2010}. In these experiment the spin effect have been neglected for modeling the signal. It is highly likely that super massive black holes have significant spin due to accretion of matter from surrounding. We have seen that the spin can significantly change memory amplitude as a result it can change the timing residuals. Hence it is important to consider the spin effect while modeling the signals. The important thing for pulsar timing measurements would be the amount of memory changes over a single two weeks period of time which is shown the figure \ref{fig:MemoryMassVsSpin} as function of total mass of the binary and the spin magnitude. For a given mass and over the allowed spin range the net change in the memory amplitude can be as big as a factor of 5 which would change the SNR, or the expected event rate within a given volume, by a factor of $5^{3}=125$

>>>>

The growth in the nonlinear memory signal amplitude from super massive black hole occurs on the time scale of a few days so the effect on timing residuals can be seen if observed over a few weeks period of time. A method for observing Nonlinear memory signal from super massive black hole mergers using pulsar timing has been proposed in ref\cite{Pshirkov2010}. The authors have shown that the present day pulsar timing precision allows one to detect BWM due to  SMBH merger at distance upto 1 Gpc (for case of equal mass $10^{8}$ solar mass). The details regarding the calculation of the timing residuals can be seen in \cite{Pshirkov2010}, where they consider the memory event as a step function which occurs in the middle of an observation time of 10 years. The method described in \cite{Pshirkov2010} is general and can be applied to case where the memory jump is not at step function. In our analysis we explore how the nonlinear memory grows during two weeks interval to time for different ranges of mass and spin parameter values. We also calculate the timing residuals as done in \cite{Pshirkov2010} and explore how it changes as a function of total mass and spin of black holes as shown in Fig[\ref{fig:MemoryMassVsSpin}].
\par 
Future observations with increased precision would be able to look at the memory growth on even shorter time of two weeks. The memory growth on the time scale of two weeks does not look like a step function, instead it's shape depends upon the spins of the merging black holes. In Fig[\ref{fig:MemoryGrowthTwoweeks}].we show the memory growth during two weeks interval of time with markers on complete memory during for different range of spin and total mass of the binary. The solid line show the complete memory growth in growth in geometrical units. It can be seen the as the total mass of binary becomes larger the two weeks interval of time shrinks in geometrical units. We have chosen this two weeks interval of time such that the change in memory during this time interval is maximum. For the pulsar timing observation we compute the residuals during this two week interval of time following the method described in \cite{Pshirkov2010}. It can be seen in Fig[\ref{fig:MemoryGrowthTwoweeks}] that the memory growth during this two weeks interval of time look very for different mass values. For higher masses ($10^{10}M_{\odot}$), this looks like a straight line. For lower masses this looks like a nonlinear growth. In Fig[\ref{fig:resedualGrowthTwoweeks}] we show the residual before(pre-fit) and after  (post-fit) a quadratic subtraction. It can be seen that the residual growth fits better to a pre-fit residuals for a higher mass supermassive black hole ($M=10^{11}M_{\odot}$), which obvious because when the memory growth is linear, it integral would give a quadratic function. However the memory growth for black holes with $M = 10^{9}M_{\odot}$ looks very  nonlinear, hence the resulting integral also looks like a function different from a quadratic polynomial. This makes the timing residual post quadratic subtraction depend upon the mass of super massive binary. Fig[\ref*{fig:resedualGrowthLogLog}] shows a log-log plot for pre-fit and post-fit residuals for different ranges of mass and spin parameters. It can be seen the residual growth amplitude when evaluated for a two weeks interval tends to increase initially with mass, reaches a maximum when $M=10^{9}M_{\odot}$ and then again decreases. This shows that when observed over a two weeks interval of time super massive black holes above a mass of $10^{10}M_{\odot}$ would be harder to detect. Fig[\ref*{fig:MemoryMassVsSpin}] show color plot for rms residuals for a range of spin and mass values.  
%%\begin{widetext}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksGeoUnits.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow8MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow9MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow11MSun.pdf}
	\caption{The top figure shows the memory growth during merger of a binary with time measured in geometric units with total spin magnitude as indicated above. The two weeks interval of time for a given mass as indicated above. The bottom three figures show the the memory growth in days for Mass $10^{8}M_{\odot}$, $10^{9}M_{\odot}$ and $10^{11}M_{\odot}$ respectively.}
	\label{fig:MemoryGrowthTwoweeks}
%\end{widetext}

@@

\begin{figure}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDays10pow8MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDays10pow8MSunQuadSub.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDays10pow10MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDays10pow10MSunQuadSub.pdf}
	\caption{The figure above shows residual growth in a given interval of time. The top left plot shows this residual growth during two weeks interval of time for different spin parameter values and the top right plot shows the residual growth after subtracting a quadratic fit for maximally spinning case and a binary with $M = 10^{8}\, M_{\odot}$. The bottom figure shows the similar plot for a binary with $M = 10^{11}\, M_{\odot}$.}
	\label{fig:resedualGrowthTwoweeks}
\end{figure}

\begin{figure}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDaysPreFit.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/ResidualGrowth2weeksDiffSpinsInDaysPostfit.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/ResidualGrowth5YearsDiffSpinsInDaysPrefit.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/ResidualGrowth5YearsDiffSpinsInDaysPostfit.pdf}
	\caption{The figure above shows residual growth in a given interval of time. The top left plot shows this residual growth during two weeks interval of time and the top right plot shows the residual growth after subtracting a quadratic fit. The bottom figure shows the similar plot for 5 years interval of time.}
	\label{fig:resedualGrowthLogLog}
\end{figure}

\begin{figure}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/MemoryResSpinvsMass14Days.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/MemoryResSpinvsMass1825Days.pdf}
	\caption{The top figure shows the root mean square of post-fit timing residuals growth as a function of Mass and Spin magnitude $\chi$. The left plot shows this post-fit residual fit during two weeks interval of time and right plot shows this for 5 years interval.}
	\label{fig:MemoryMassVsSpin}
\end{figure}
\end{widetext}
\section{CONCLUSIONS}
Observation of Christodoulou memory would be an important test of general relativity in strong field regime. Effort to detect this memory effect in low frequency band from super massive black holes have been undertaken by pulsar timing array experiments \cite{Arzoumanian2015}. The author in their analysis have looked for memory in a observation data for 5 years. In their analysis they have assumed the memory to a step function for a given amplitude. This amplitude is of the order $10^{-15}$, which is memory jump for a $10^{8}M_{\odot}$ binary. In these analysis the spin effect have been neglected for modeling the signal. It is highly likely that super massive black holes have significant spin due to accretion of matter from surrounding. We have seen that the spin can significantly change memory amplitude as a result it can change the timing residuals. In our analysis we observe that:(1) Growth of memory when looks over a shorter observing time scale does not look like a step function, instead it's shape depends upon the spin and mass of binary for a give interval of time.(2) The post-fit residual for a observation period of two weeks  .Hence it is important to consider the spin effect while modeling the signals. The important thing for pulsar timing measurements would be the amount of memory changes over a single two weeks period of time which is shown the figure \ref{fig:MemoryMassVsSpin} as function of total mass of the binary and the spin magnitude. For a given mass and over the allowed spin range the net change in the memory amplitude can be as big as a factor of 5 which would change the SNR, or the expected event rate within a given volume, by a factor of $5^{3}=125$
